---
title: "Presentation of the mixedML package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth : 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Presentation of the mixedML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
devtools::load_all()
```


# Introduction

This package provides functions to train hybrid mixed effects models. Such models are a variation of linear mixed effects models, used for Gaussian longitudinal data, whose formulation is:

$$Y_{ij} = X_{ij} \beta +  Z_{ij} u_i + w_{i} + \varepsilon_{ij}$$

… where:

- $i$ is the subject,
- $j$ is the observation,
- $\beta$ is the vector of fixed effects and $X_{ij}$ is associated covariates,
- $u_i$ is the vector of random effects and $Z_{ij}$ the associated covariates,
- $w_i$ is a zero-mean Gaussian stochastic process (such as Brownian motion) modeling the correlation in the individual error,
- $\epsilon_{ij}$ is the zero-mean Gaussian residual error.
-

For such hybrid models:

-   a Machine Leaning (ML) model is used to estimates the fixed effects;
-   a Mixed Effects model (`hlme` from [lcmm package](https://cecileproust-lima.github.io/lcmm/articles/lcmm.html)) is constrained to estimate only random effects.

That is, the formulation becomes:

$$Y_{ij} = f_{ML}(X_{ij}) +  Z_{ij} u_i + w_{i} + \varepsilon_{ij}$$


… where $f_{ML}(X_{ij})$ is the output from a ML model trained to predict the fixed effects.

Using ML models to estimates the fixed effects has two main advantages comparing to linear models:

-   they can handle highly non-linear relations, and do so with simple inputs (instead of being highly dependent of the specification);
-   they can handle complex time interactions, in the case of Recurrent Neural Networks;

However, some ML models have a "black box" effect, as one cannot use its estimated parameters to understand the relations within the data.

# Method

The method uses a iterative training of both fixed effects and random effects models.

So far we are fitting each model on the **residuals** of the other.
This is only valid for **regression problems** where

- in generalized linear model terms: the models use an identity link,
- in neural networks terms: the models do not use a final activation function.

In such case, fitting on the residuals is equivalent to fitting using the predictions (of the other model) as offset.


The pseudo-code is as follow (`fe`/`re` stands for fixed/random effects):
```{r, eval=FALSE}
ml_model_fe <- initiate_ml_model_fe()
hlme_model_re <- initiate_hlme_model_re()

Yre <- 0.
while not converged:
  fit ml_model_fe on X and (Y - Yre)
  Yfe <-  ml_model_fe(X)

  fit hlme_model_re on X and (Y - Yfe)
  Yre <-  hlme_model_re(X)

  converged <- criterion(Y, Yfe+Yre)
```

This method works with any kind of model, so far models coded in R and Python acn be easily implemented.


# Example dataset

The dataset `data_mixedml` is proposed. It is generated using the `R/data_gen.R` file.

It is a synthetic longitudinal dataset, containing data for 10 subjects on 5 regularly spaced time steps.
It contains two response columns for both fixed and mixed effects.
NA values have been added manually, and some observations are deleted.
Also the row names have been shuffled in order to enforce a clear use of row names or indices in the code.

```{r}
data_mixedml
```


# Main/fit functions

## Formalism

The MixedML models are obtained using specific functions which have for signature:

```{r, eval=FALSE}
XXXX_mixedml(
  # parameters of the MixedML model (inpired by the hlme function definition)
  fixed_spec,
  random_spec,
  data_,
  subject,
  time,
  # parameters for MixedML method
  mixedml_controls,
  # controls (extra-parameters) for the hlme model
  hlme_controls,
  # controls (extra-parameters) specific to the implemented ML model
  controls_1, controls_2, et_caetera
)
```


… where `XXXX` takes the name of the ML model used for the fixed effects. As an example, the Reservoir Computing model
is named `reservoir_mixedml`.

Using a dedicated function allows to benefit from the code-completion since every arguments is explicitely specified
(no optional arguments).


## Arguments

The `fixed_spec`, `random_spec`, `cor`, `data`, `subject` and `time` are used by both sub-models and are taken
from the `hlme` function which can be seen in
[the lcmm package documentation](https://cecileproust-lima.github.io/lcmm/reference/hlme.html)

Then several controls are defined, using specific functions whose names correspond to the control names.
That is, the `some_name_ctrls(…)` function is used to define `some_name_controls` controls.
Each control has its specific help.

## Example
Here is an example using the `reservoir_mixedml` function (here is the [corresponding vignette](mixedML_reservoir.html)):

```{r}
data_train <- data_mixedml[data_mixedml$ID < 9, ]
data_val <- data_mixedml[data_mixedml$ID == 9, ]
data_test <- data_mixedml[data_mixedml$ID == 10, ]

model_reservoir <- reservoir_mixedml(
  fixed_spec = ym ~ 1 + x1 + x2 + x3,
  random_spec = ~ 1 + x1 + x2,
  data = data_train,
  data_val = data_val,
  subject = "ID",
  time = "time",
  # parameters for MixedML method
  mixedml_controls = mixedml_ctrls(
    earlystopping_controls = earlystopping_ctrls(min_mse_gain = 0.1, patience = 3),
    all_info_hlme_prediction = TRUE
  ),
  # controls (extra-parameters) for the hlme model
  hlme_controls = hlme_ctrls(maxiter = 50, idiag = TRUE),
  # controls (extra-parameters) for the ML model
  esn_controls = esn_ctrls(units = 20, ridge = 1e-5),
  ensemble_controls = ensemble_ctrls(seed_list = c(1, 2, 3, 4, 5)),
  fit_controls = fit_ctrls(warmup = 1)
)
```

The resulting model will be used in the remaining sections.


# Model attributes

Each sub-models are accessible from the fitted MixedML model:

```{r}
model_reservoir$random_model
```

```{r}
model_reservoir$fixed_model
```

Also a `call` attribute exists, meaning one can trained the model with new inputs using `update` command:

```{r, eval=FALSE}
new_model_reservoir <- update(model_reservoir, data = new_data, maxiter = new_maxiter)
```

# Post-fit functions

The following functions are common to all fitted MixedML models.


## `summary`

```{r}
summary(model_reservoir)
```


## `plot_convergence`

```{r, fig.width=10}
plot_convergence(model = model_reservoir, ylog_mse = TRUE)
```

## `plot_prediction_check`


```{r, fig.width=10, fig.height=20}
plot_predictions_check(model = model_reservoir)
```



## `predict`

```{r}
pred_test_past <- predict(
  model = model_reservoir,
  data = data_test,
  all_info_hlme_prediction = FALSE,
  nproc_hlme_past = 1
)
pred_test_past
```


```{r}
pred_test_all <- predict(
  model = model_reservoir,
  data = data_test,
  all_info_hlme_prediction = TRUE,
  nproc_hlme_past = 1
)
pred_test_all
```


## `plot_predictions`

```{r}
hlme_model <- lcmm::hlme(
  fixed = ym ~ 1 + x1 + x2 + x3,
  random = ~ 1 + x1 + x2,
  data = data_train,
  subject = "ID",
  var.time = "time"
)
hlme_preds <- lcmm::predictY(hlme_model, newdata = data_test)
hlme_preds_all <- hlme_preds$pred
names(hlme_preds_all) <- rownames(hlme_preds$times)

plot_predictions(
  model = model_reservoir,
  data_pred = data_test,
  list_preds = list(
    mixedml_all_info = pred_test_all,
    mixedml_past_info = pred_test_past,
    hlme_all_info = hlme_preds_all
  ),
  ncols = 1
)
```



## `save_mixedml`

This function is used to save a mixedML model. It is **mandatory** when using a model based on a Python package, since
we need to save both R and Python objects.


```{r}
save_mixedml(model_reservoir, mixedml_model_rds = "model_reservoir.Rds")
```


## `load_mixedml`

This function is used to load a mixedML model. It is **mandatory** when using a model based on a Python package, since
we need to load both R and Python objects.

```{r}
mixedml_model <- load_mixedml("model_reservoir.Rds")
```

```{r}
mixedml_model$fixed_model
```

```{r}
mixedml_model$random_model
```


# Remark on logging

The use of reticulate makes it cumbersome to implement logging in the package. Since the solutions found involve
preventing the use of Rstudio and favor running from terminal, one simple solution is to write a standard R script and
call it from the terminal, redirecting both stdout and stderr to a log file:

```bash
Rscript name_of_script.R > log_file.log 2>&1
```



```{r, include=FALSE}
unlink("model_reservoir.Rds")
unlink("model_reservoir.Rds.joblib")
```
