---
title: "Presentation of the mixedML package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth : 3
    number_sections: true
    includes:
    css: vignette-style.css
vignette: >
  %\VignetteIndexEntry{Presentation of the mixedML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
devtools::load_all()
source("utils.R")
```


# Introduction

This package provides functions to train hybrid mixed effects models. Such models are a variation of linear mixed effects models, used for Gaussian longitudinal data, whose formulation is:

$$Y_{ij} = X_{ij} \beta +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$

… where $i$ is the subject, $j$ is the occasion, and $w_i$ comes from a zero-mean Gaussian stochastic process (such as Brownian motion).

<br><br> For such hybrid models:

-   a Machine Leaning (ML) model is used to estimates the fixed effects;
-   a Mixed Effects model (`hlme` from [lcmm package](https://cecileproust-lima.github.io/lcmm/articles/lcmm.html)) is constrained to estimate only random effects.

That is, the formulation becomes:

$$Y_{ij} = f_{ML}(X_{ij}) +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$


… where $f_{ML}(X_{ij})$ is the output from a ML model trained to predict the fixed effects.

<br><br> Using ML models to estimates the fixed effects has two main advantages comparing to linear models:

-   they can handle highly non-linear relations, and do so with simple inputs (instead of being highly dependent of the specification);
-   they can handle complex time interactions, in the case of Recurrent Neural Networks;

However, some ML models have a "black box" effect, as one cannot use its estimated parameters to understand the relations within the data.

# Method

The method uses a iterative training of both fixed effects and random effects models.

So far we are fitting each model on the **residuals** of the other.
This is only valid for **regression problems** where it is equivalent to
to fitting using the predictions (of the other model) as offset
(see [this vignette](method_study.html) for an example).
In generalized linear model terms: the models use an identity link.
In neural networks terms: the models do not use a final activation function.


The pseudo-code is as follow (`fe`/`re` stands for fixed/random effects):
```
ml_model_fe <- initiate_ml_model_fe()
hlme_model_re <- initiate_hlme_model_re()

Yre <- 0.
while not converged:
  fit ml_model_fe on X and (Y - Yre)
  Yfe <-  ml_model_fe(X)

  fit hlme_model_re on X and (Y - Yfe)
  Yre <-  hlme_model_re(X)

  converged <- criterion(Y, Yfe+Yre)
```

# Example dataset

The dataset `data_mixedml` is proposed. It is generated using the `R/data_gen.R` file.

It is a synthetic longitudinal dataset, containing data for 10 subjects on 5 regularly spaced time steps. It contains two response columns for both fixed and mixed effects. NA values have been added manually in these columns:

```{r}
data_mixedml
```


# General principle

The MixedML models are obtained using specific functions which have for signature:

```{r, eval=FALSE}
some_mixed_ml_model(
  # parameters of the MixedML model (inpired by the hlme function definition)
  fixed_spec,
  random_spec,
  data_,
  subject,
  time,
  # parameters for MixedML method
  mixedml_controls,
  # controls (extra-parameters) for the hlme model
  hlme_controls,
  # controls (extra-parameters) for the implemented ML model
  controls_1, controls_2, et_caetera
)
```

# Arguments

The `fixed_spec`, `random_spec`, `cor`, `data`, `subject` and `time` are used by both sub-models and are taken from the `hlme` function which can be seen in [the lcmm package documentation](https://cecileproust-lima.github.io/lcmm/reference/hlme.html)

Then several controls are defined, using specific functions whose names correspond to the control names. That is, the `some_name_ctrls(…)` function is used to define `some_name_controls` controls. Each control has its specific help.

# Example
Here is an example using the `reservoir_mixedml` function (here is the [corresponding vignette](mixedML_reservoir.html):

```{r}
data_train <- data_mixedml[data_mixedml$ID < 9, ]
data_val <- data_mixedml[data_mixedml$ID >= 9, ]

model_reservoir <- reservoir_mixedml(
  fixed_spec = ym ~ x1 + x2 + x3,
  random_spec = ~ x1 + x2,
  data = data_train,
  data_val = data_val,
  subject = "ID",
  time = "time",
  # parameters for MixedML method
  mixedml_controls = mixedml_ctrls(
    earlystopping_controls = earlystopping_ctrls(min_mse_gain = 0.1, patience = 10),
    all_info_hlme_prediction = TRUE
  ),
  # controls (extra-parameters) for the hlme model
  hlme_controls = hlme_ctrls(maxiter = 50, idiag = TRUE),
  # controls (extra-parameters) for the ML model
  esn_controls = esn_ctrls(units = 20, ridge = 1e-5),
  ensemble_controls = ensemble_ctrls(seed_list = c(1, 2, 3, 4, 5)),
  fit_controls = fit_ctrls(warmup = 1)
)
```

The resulting model will be used in the remaining sections.


# Remark on logging

The use of reticulate makes it cumbersome to implement logging in the package. Since the solutions found involve
preventing the use of Rstudio and favor running from terminal, one simple solution is to write a standard R script and
call it from the terminal, redirecting both stdout and stderr to a log file:

```bash
Rscript name_of_script.R > log_file.log 2>&1
```

# Remark on Python environments

One might to use a specific Python environment for the reticulate part of the package.

To do so you have two options:

- Using the `use_python_environment` function from the package.
- Setting the environement variable `MIXED_ML_PYTHON_ENV` (which will be automatically read at package load time).

The value of the variable should be set as `conda:environement_name` or `virtualenv:environement_name` depending on
the type of environment you want to use.


# Attributes

Each sub-models are accessible from the fitted MixedML model:

```{r}
model_reservoir$random_model
```

```{r}
# (this model uses reticulate so it not very convenient as an example…)
model_reservoir$fixed_model
```

Also a `call` attribute exists, meaning one can trained the model with new inputs using `update` command:

```{r, eval=FALSE}
new_model_reservoir <- update(model_reservoir, data = new_data, maxiter = new_maxiter)
```

# Functions

The function `predict`, `plot_conv`, `plot_best_iter` are common to all fitted MixedML models.

The function `load_backup` can be used to inspect the model and the predictions of a specific iteration.

## `predict`

```{r, echo=FALSE}
print_help("predict")
```


```{r}
predict(
  model = model_reservoir,
  data = data_mixedml,
  all_info_hlme_prediction = FALSE
)
```



## `plot_conv`

```{r, echo=FALSE}
print_help("plot_conv_mse")
```

```{r}
plot_conv_mse(model = model_reservoir, ylog = TRUE)
```


## `plot_loglik`

```{r, echo=FALSE}
print_help("plot_conv_loglik")
```

```{r}
plot_conv_loglik(model = model_reservoir)
```

## `plot_prediction_check`

```{r, echo=FALSE}
print_help("plot_prediction_check")
```

```{r}
plot_prediction_check(model = model_reservoir, subject_nb_or_list = c(1, 2, 3, 4, 5))
```


## `save_mixedml`

```{r, echo=FALSE}
print_help("save_mixedml")
```


```{r}
save_mixedml(model_reservoir, mixedml_model_rds = "model_reservoir.Rds")
```


## `load_mixedml`


```{r, echo=FALSE}
print_help("load_mixedml")
```

```{r}
mixedml_model <- load_mixedml("model_reservoir.Rds")
```

```{r}
mixedml_model$fixed_model
```

```{r}
mixedml_model$random_model
```

```{r}
predict(mixedml_model, data_mixedml)
```



```{r, echo=FALSE}
unlink("model_reservoir.Rds")
unlink("model_reservoir.Rds.joblib")
```
