---
title: "Presentation of the mixedML package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth : 3
    number_sections: true
    includes:
    css: vignette-style.css
vignette: >
  %\VignetteIndexEntry{Presentation of the mixedML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(digits = 3)
devtools::load_all()
source("utils.R")
```


# Introduction

This package provides functions to train hybrid mixed effects models. Such models are a variation of linear mixed effects models, used for Gaussian longitudinal data, whose formulation is:

$$Y_{ij} = X_{ij} \beta +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$

… where $i$ is the subject, $j$ is the occasion, and $w_i$ comes from a zero-mean Gaussian stochastic process (such as Brownian motion).

<br><br> For such hybrid models:

-   a Machine Leaning (ML) model is used to estimates the fixed effects;
-   a Mixed Effects model (`hlme` from [lcmm package](https://cecileproust-lima.github.io/lcmm/articles/lcmm.html)) is constrained to estimate only random effects.

That is, the formulation becomes:

$$Y_{ij} = f_{ML}(X_{ij}) +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$


… where $f_{ML}(X_{ij})$ is the output from a ML model trained to predict the fixed effects.

<br><br> Using ML models to estimates the fixed effects has two main advantages comparing to linear models:

-   they can handle highly non-linear relations, and do so with simple inputs (instead of being highly dependent of the specification);
-   they can handle complex time interactions, in the case of Recurrent Neural Networks;

However, some ML models have a "black box" effect, as one cannot use its estimated parameters to understand the relations within the data.

# Method

The method uses a iterative training of both fixed effects and random effects models.

So far we are fitting each model on the residuals of the other.
This is only valid for **regression problems** where it is equivalent to
to fitting using the predictions (of the other model) as offset.
In generalized linear model terms: the models use an identity link.
In neural networks terms: the models do not use a final activation function.


The pseudo-code is as follow (`fe`/`re` stands for fixed/random effects):
```
ml_model_fe <- initiate_ml_model_fe()
hlme_model_re <- initiate_hlme_model_re()

Yre <- 0.
while not converged:
  fit ml_model_fe on X and (Y - Yre)
  Yfe <-  ml_model_fe(X)

  fit hlme_model_re on X and (Y - Yfe)
  Yre <-  hlme_model_re(X)

  converged <- criterion(Y, Yfe+Yre)
```

# Example dataset

The dataset `data_mixedml` is proposed. It is generated using the `R/data_gen.R` file.

It is a synthetic longitudinal dataset, containing data for 10 subjects on 5 regularly spaced time steps. It contains two response columns for both fixed and mixed effects. NA values have been added manually in these columns:

```{r}
data_mixedml
```


# General principle

The MixedML models are obtained using specific functions which have for signature:

```{r, eval=FALSE}
some_mixed_ml_model(
  # parameters of the MixedML model (inpired by the hlme function definition)
  fixed_spec,
  random_spec,
  data_,
  subject,
  time,
  # parameters for MixedML method
  mixedml_controls,
  # controls (extra-parameters) for the hlme model
  hlme_controls,
  # controls (extra-parameters) for the implemented ML model
  controls_1, controls_2, et_caetera
)
```

# Arguments

The `fixed_spec`, `random_spec`, `cor`, `data`, `subject` and `time` are used by both sub-models and are taken from the `hlme` function which can be seen in [the lcmm package documentation](https://cecileproust-lima.github.io/lcmm/reference/hlme.html)

Then several controls are defined, using specific functions whose names correspond to the control names. That is, the `some_name_ctrls(…)` function is used to define `some_name_controls` controls. Each control has its specific help.

# Example
Here is an example using the `reservoir_mixedml` function (here is the [corresponding vignette](mixedML_reservoir.html):

```{r}
model_reservoir <- reservoir_mixedml(
  fixed_spec = ym ~ x1 + x2 + x3,
  random_spec = ~ x1 + x2,
  data = data_mixedml,
  subject = "ID",
  time = "time",
  # parameters for MixedML method
  mixedml_controls = mixedml_ctrls(),
  # controls (extra-parameters) for the hlme model
  hlme_controls = hlme_ctrls(maxiter = 50, idiag = TRUE),
  # controls (extra-parameters) for the ML model
  esn_controls = esn_ctrls(units = 20, ridge = 1e-5),
  ensemble_controls = ensemble_ctrls(seed_list = c(1, 2, 3, 4, 5)),
  fit_controls = fit_ctrls(warmup = 2),
  output_dir = "mixedML_vignette"
)
```

The resulting model will be used in the remaining sections.

# Attributes

Each sub-models are accessible from the fitted MixedML model:

```{r}
model_reservoir$random_model
```

```{r}
# (this model uses reticulate so it not very convenient as an example…)
model_reservoir$fixed_model
```

Also a `call` attribute exists, meaning one can trained the model with new inputs using `update` command:

```{r, eval=FALSE}
new_model_reservoir <- update(model_reservoir, data = new_data, maxiter = new_maxiter)
```

# Functions

The function `predict`, `plot_conv`, `plot_last_iter` are common to all fitted MixedML models.

The function `load_backup` can be used to inspect the model and the predictions of a specific iteration.

## `predict`

```{r, echo=FALSE}
print_help("predict")
```

## `plot_conv`

```{r, echo=FALSE}
print_help("plot_conv")
```

```{r}
plot_conv(model = model_reservoir)
```


## `plot_loglik`

```{r, echo=FALSE}
print_help("plot_loglik")
```

```{r}
plot_loglik(model = model_reservoir)
```

## `plot_last_iter`

```{r, echo=FALSE}
print_help("plot_last_iter")
```

```{r}
plot_last_iter(model = model_reservoir, subject_nb_or_list = c(1, 2, 3, 4, 5))
```


## `load_backup`


```{r}
backup <- load_backup(
  fixed_model_rds_or_joblib = "mixedML_vignette/007_fixed_model.joblib",
  random_model_rds = "mixedML_vignette/007_random_model.Rds"
)
```

```{r}
backup$fixed_model
```

```{r}
backup$random_model
```

```{r}
predict(backup, data_mixedml)
```




```{r}
unlink("mixedML_vignette", recursive = TRUE)
```
