---
title: "Presentation of the mixedML package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth : 3
    number_sections: true
    includes:
    css: vignette-style.css
vignette: >
  %\VignetteIndexEntry{Presentation of the mixedML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
devtools::load_all()
```

```{r, include=FALSE}
clean_help_md <- function(help_md) {
  help_md <- gsub(pattern = "^# .*\n", replacement = "", help_md, perl = TRUE)
  help_md <- gsub("## (.*)\n", "**\\1**\n", help_md, perl = TRUE)
  help_md <- gsub("\n+$", "", help_md, perl = TRUE)
  return(help_md)
}


print_help <- function(function_name) {
  stopifnot(is.character(function_name))
  hpath <- as.character(help(function_name))[[3]]
  markdown_doc <- Rd2md::as_markdown(Rd2md::read_rdfile(hpath))
  markdown_doc <- clean_help_md(markdown_doc)
  return(knitr::asis_output(markdown_doc))
}
```

# Introduction

This package provides functions to train hybrid mixed effects models. Such models are a variation of linear mixed effects models, used for Gaussian longitudinal data, whose formulation is:

$$Y_{ij} = X_{ij} \beta +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$

… where $i$ is the subject, $j$ is the occasion, and $w_i$ comes from a zero-mean Gaussian stochastic process (such as Brownian motion).

<br><br> For such hybrid models:

-   a Machine Leaning (ML) model is used to estimates the fixed effects;
-   a Mixed Effects model (`hlme` from [lcmm package](https://cecileproust-lima.github.io/lcmm/articles/lcmm.html)) is constrained to estimate only random effects.

That is, the formulation becomes:

$$Y_{ij} = f_{ML}(X_{ij}) +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$$

… where $f_{ML}(X_{ij})$ is the output from a ML model trained to predict the fixed effects.

<br><br> Using ML models to estimates the fixed effects has two main advantages comparing to linear models:

-   they can handle highly non-linear relations, and do so with simple inputs (instead of being highly dependent of the specification);
-   they can handle complex time interactions, in the case of Recurrent Neural Networks;

However, some ML models have a "black box" effect, as one cannot use its estimated parameters to understand the relations within the data.

# Method

The method uses a iterative training of both fixed effects and random effects models. The pseudo-code is as follow (`fe`/`re` stands for fixed/random effects):

```         
ml_model_fe <- initiate_ml_model_fe()
hlme_model_re <- initiate_hlme_model_re()

Yre <- 0.
while not converged:
  fit ml_model_fe on X and (Y - Yre)
  Yfe <-  ml_model_fe(X)

  fit hlme_model_re on X and (Y - Yfe)
  Yre <-  hlme_model_re(X)
  
  converged <- criterion(Y, Yfe+Yre)
```

# Example dataset

The dataset `data_mixedml` is proposed: it is a synthetic longitudinal dataset, containing data for 500 subjects on 26 regularly spaced time steps ($t \in [0,25]$).

Here's a snippet:

```{r}
idx_print <- (data_mixedml[["subject"]] < 4) & (data_mixedml[["time"]] < 5)
data_print <- data_mixedml[idx_print,]
row.names(data_print) <- NULL
print(data_print, digits = 2, row.names = FALSE)
```

<br> The purely fixed effects response, $y_{fixed}$ is calculated as:

$$y_{fixed} = \gamma_{0} + 
\gamma_{1} \cdot  x_2 \cdot x_5 + 
\gamma_{2} \cdot  x_4 \cdot x_7 +
\gamma_{3} \cdot  x_6 \cdot x_8$$

… where:

- $\gamma_{0}= -0.9826036$, 
- $\gamma_{1} = -0.4289147$, 
- $\gamma_{2} = -0.0456483$,
- $\gamma_{3} = -0.8542527$.

<br> The mixed effects response, $y_{mixed}$ is calculated for each individual $i$ as: 

$$y_{mixed,i} = \gamma_{0,i} + 
\gamma_{1,i} \cdot  x_2 \cdot x_5 + 
\gamma_{2,i} \cdot  x_4 \cdot x_7 +
\gamma_{3,i} \cdot  x_6 \cdot x_8$$ 

…where each $\gamma_{k,i}$ is draw from a normal distribution with: 

- for $\gamma_{0,i}$, a mean of $\gamma_{0}$ and a standard deviation of $0.5$,
- for $\gamma_{1,i}$, a mean of $\gamma_{1}$ and a standard deviation of $0.5$, 
- for $\gamma_{2,i}$, a mean of $\gamma_{2}$ and a standard deviation of $0.05$, 
- for $\gamma_{3,i}$, a mean of $\gamma_{3}$ and a standard deviation of $0.1$.

That is, one can train a MixedML model using $y_mixed$ then use $y_fixed$ to check how well both sub-model are fitted for their specific task.

Let's defined sub-datasets for the next examples:

```{r}
data_1 <- data_mixedml[data_mixedml[["subject"]] %in% seq(01, 10), ]
data_2 <- data_mixedml[data_mixedml[["subject"]] %in% seq(11, 12), ]
```

# General principle

The MixedML models are obtained using specific functions which have for signature:

```{r, eval=FALSE}
some_mixed_ml_model(
  # parameters of the MixedML model (inpired by the hlme function definition)
  fixed_spec,
  random_spec,
  data,
  subject,
  time,
  # parameters for MixedML method
  mixedml_controls,
  # controls (extra-parameters) for the hlme model
  hlme_controls,
  # controls (extra-parameters) for the implemented ML model
  controls_1, controls_2, et_caetera
)
```

# Arguments

The `fixed_spec`, `random_spec`, `cor`, `data`, `subject` and `time` are used by both sub-models and are taken from the `hlme` function which can be seen in [the lcmm package documentation](https://cecileproust-lima.github.io/lcmm/reference/hlme.html)

Then several controls are defined, using specific functions whose names correspond to the control names. That is, the `some_name_ctrls(…)` function is used to define `some_name_controls` controls. Each control has its specific help.

Here is an example using the `reservoir_mixedml` function (the resulting model will be used in the remaining sections):

```{r}
model_reservoir <- reservoir_mixedml(
  fixed_spec = y_mixed ~ x1 + x2 + x3 + x8,
  random_spec = y_mixed ~ x1 + x2 + x3 + x8,
  data = data_1,
  subject = "subject",
  time = "time",
  # parameters for MixedML method
  mixedml_controls = mixedml_ctrls(),
  # controls (extra-parameters) for the hlme model
  hlme_controls = hlme_ctrls(nproc = 5, maxiter = 10, idiag = TRUE),
  # controls (extra-parameters) for the ML model
  esn_controls = esn_ctrls(units = 20, ridge = 1e-5),
  ensemble_controls = ensemble_ctrls(seed_list = c(1, 2, 3)),
  fit_controls = fit_ctrls(warmup = 2)
)
```

## Attributes

Each sub-models are accessible from the fitted MixedML model:

```{r}
model_reservoir$random_model
```

```{r}
# (this model uses reticulate so it not very convenient as an example…)
model_reservoir$fixed_model
```

Also a `call` attribute exists, meaning one can trained the model with new inputs using `update` command:

```{r, eval=TRUE}
updated_model <- update(model_reservoir, data = data_2)
```

# Functions

The function `predict`, `plot_conv` and `plot_last_iter` are common to all fitted MixedML models.

## `predict`

```{r, echo=FALSE}
print_help("predict")
```

## `plot_conv`

```{r, echo=FALSE}
print_help("plot_conv")
```

```{r}
plot_conv(model = model_reservoir)
```

## `plot_last_iter`

```{r, echo=FALSE}
print_help("plot_last_iter")
```

```{r, echo=FALSE}
plot_last_iter(model = model_reservoir, subject_nb_or_list = 3)
```
