---
title: "Presentation of the mixedML package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth : 3
vignette: >
  %\VignetteIndexEntry{Presentation of the mixedML package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
devtools::load_all()
```

```{r, include=FALSE}
clean_help_md <- function(help_md) {
  help_md <- gsub(pattern = "^# .*\n", replacement = "", help_md, perl = TRUE)
  help_md <- gsub("## (.*)\n", "**\\1**\n", help_md, perl = TRUE)
  return(help_md)
}


print_help <- function(function_name) {
  stopifnot(is.character(function_name))
  hpath <- as.character(help(function_name))[[3]]
  markdown_doc <- Rd2md::as_markdown(Rd2md::read_rdfile(hpath))
  markdown_doc <- clean_help_md(markdown_doc)
  knitr::asis_output(markdown_doc)
  return()
}
```

# Introduction

This package provides functions to train hybrid mixed effects models. Such models are a variation of linear mixed effects models, used for Gaussian longitudinal data, whose formulation is:

$Y_{ij} = X_{ij} \beta +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$

… where $i$ is the subject, $j$ is the occasion, and $w_i$ comes from a zero-mean Gaussian stochastic process (such as Brownian motion),


For such hybrid models:

- a Machine Leaning (ML) model is used to estimates the fixed effects;
- a Mixed Effects model (`hlme` from [lcmm package](https://cecileproust-lima.github.io/lcmm/articles/lcmm.html)) is used to estimate the random effects.

That is, the formulation becomes:

$Y_{ij} = f_{ML}(X_{ij}) +  Z_{ij} u_i + w_{ij} + \varepsilon_{ij}$

… where $f_{ML}(X_{ij})$ is the output from a ML model trained to predict the fixed effects.


Using ML models to estimates the fixed effects have tow main advantages comparing to linear models:

- they can handle highly non-linear relations, and do so with simple inputs (instead of being highly dependent of the specification);
- they can handle complex time interactions, in the case of Recurrent Neural Networks;

However, some ML models have a "black box" effect, as one cannot use its estimated parameters to understand the relations within the data.

# General MixedML use

## General principle

The MixedML models are obtained using specific functions which have for signature:

```{r, eval=FALSE}
some_mixed_ml_model(
  # parameters of the MixedML model (inpired by the hlme function definition)
  fixed_spec,
  random_spec,
  data,
  subject,
  time,
  # parameters for MixedML method
  mixedml_controls,
  # controls (extra-parameters) for the hlme model
  hlme_controls,
  # controls (extra-parameters) for the ML model
  controls_1, controls_2, et_caetera
)
```

The `fixed_spec`, `random_spec`, `cor`, `data`, `subject` and `time` parameters are taken from the `hlme` function and can be seen in [the lcmm package documentation](https://cecileproust-lima.github.io/lcmm/reference/hlme.html)

## Controls

Controls are defined using specific functions, whose names correspond to the control names: the `some_name_ctrls` function is used to define `some_name_controls` controls.

`mixedml_controls` and `hlme_controls` are common to all MixedML models.

### `mixedml_controls`

```{r, echo=FALSE}
print_help("mixedml_ctrls")
```

### `hlme_controls`

```{r, echo=FALSE}
print_help("hlme_ctrls")
```

## Functions

The function `predict` and `plot_conv` are common to all the MixedML models

### `predict`

```{r, echo=FALSE}
print_help("predict")
```

### `plot_conv`

```{r, echo=FALSE}
print_help("plot_conv")
```

# Implementations

So far, a hybrid model based on Reservoir Computing is available. They are implemented by interfacing the [reservoirpy](https://github.com/reservoirpy/reservoirpy) Python package with R using [reticulate](https://github.com/rstudio/reticulate).


## MixedML with Reservoir Computing

The function `reservoir_mixedml` is used to define and fit a MixedML model which uses an Reservoir Computing to fit the fixed effect.

```{r, echo=FALSE}
print_help("reservoir_mixedml")
```

Reservoir Computing is implemented using an ensemble of Echo State Network (Reservoir + Ridge readout), whose Reservoirs are initialized with different random seeds. The prediction of the ensemble model is calculated as the mean or median of all the ENS prediction, which reduces the impact of the Reservoir initialization on the results.

Four controls are used to define the RC model's behavior.

### `esn_controls`

```{r, echo=FALSE}
print_help("esn_ctrls")
```

### `ensemble_controls`

```{r, echo=FALSE}
print_help("ensemble_ctrls")
```

## `fit_controls`

```{r, echo=FALSE}
print_help("fit_ctrls")
```


# Example

```{r}
model <- reservoir_mixedml(
  fixed_spec = y_mixed ~ x1 + x2 + x3 + x8,
  random_spec = y_mixed ~ x1 + x2 + x3 + x8,
  data = data_mixedml[data_mixedml["subject"] < 10, ],
  subject = "subject",
  time = "time",
  mixedml_controls = mixedml_ctrls(),
  hlme_controls = hlme_ctrls(nproc = 5, maxiter = 10, idiag = TRUE),
  esn_controls = esn_ctrls(units = 20, ridge = 1e-5),
  ensemble_controls = ensemble_ctrls(seed_list = c(1, 2, 3)),
  fit_controls = fit_ctrls(warmup = 2)
)
```

```{r}
summary(model$random_model)
```

```{r}
plot_conv(model)
```



```{r}
plot_last_iter(model, subject_nb_or_list = 5)
```
